### 2017-10-16 Draft2.tex
More madness

	Adding is the most common operation done on dynamic arrays\footnote{}. Funnel arrays improve the performance of adding in two ways: by allocating less memory, and reducing the amount of copying. We begin with the implementation for dynamic arrays.

@@

	[proof]
	The capacity starts out at zero and becomes $\initcapacity$ once one element is added. From then on, it can only grow by a factor of $\growthfactor$ via [resize], so if $\growthfactor^i\initcapacity$ is the current capacity then $\growthfactor^{i + 1}\initcapacity$ must be the next capacity. By induction, the rest of the sequence is $$\left\{ \growthfactor^i\initcapacity \right\}_{i = 1}^k$$ for some $k$. Since the capacity is only as big as is needs to be, namely greater than or equal to $\varnitems$, $k$ is the smallest integer for which $\growthfactor^k\initcapacity \geq \varnitems$. Using this property, we now derive $k$.

@@

	[make lemma]
	Now, consider that to grow by $\growthfactor$ we have to call [resize]. So $\growthfactor^k\initcapacity$ being present in $S_C$ means we once called [resize] at size $\growthfactor^{k - 1}\initcapacity$ for $k \geq 1$. Then we must have called [resize] at the sizes $$S_R = 0,\ \initcapacity,\ \growthfactor\initcapacity,\ \growthfactor^2\initcapacity,\ \ldots\ \growthfactor^{\lceil \log_{\growthfactor} \varnitems - \log_{\growthfactor} \initcapacity \rceil - 1}\initcapacity$$

	Each time [resize] is called, we copy $i$ items where $i$ is the current size of the dynamic array. Then the total number of items copied for $\varnitems$ adds is

@@

	To make this expression easier to work with, I will write it in big-P notation to approximate it for large $\varnitems$. First I note that $\log_{\growthfactor} k \leq \lceil \log_{\growthfactor} k \rceil < \left( \log_{\growthfactor} k \right) + 1$ for any $k > 0$, so $k \leq \growthfactor^{\lceil \log_{\growthfactor} k \rceil} < \growthfactor{k}$ and $\exists r \in {[1, \growthfactor)} : rk = \growthfactor^{\lceil \log_{\growthfactor} k \rceil}$. Then

	\begin{align*}
	\nwritesfn(\varnitems) &= \varnitems + \left( \frac{\growthfactor^{\big\lceil \log_{\growthfactor} \left( \frac {\varnitems} {\initcapacity} \right) \big\rceil} - 1}{\growthfactor - 1} \right) \initcapacity\\
	\varnitems + \left( \frac{\frac{\varnitems}{\initcapacity} - 1}{\growthfactor - 1} \right) \initcapacity \leq \nwritesfn(\varnitems) &< \varnitems + \left( \frac{\growthfactor\frac{\varnitems}{\initcapacity} - 1}{\growthfactor - 1} \right) \initcapacity\\
	\varnitems + \left( \frac{\varnitems - \initcapacity}{\growthfactor - 1} \right) \leq \nwritesfn(\varnitems) &< \varnitems + \left( \frac{\growthfactor\varnitems - \initcapacity}{\growthfactor - 1} \right)\\
	\biggo\left( \varnitems + \left( \frac{\varnitems - \initcapacity}{\growthfactor - 1} \right) \right) \leq \biggo\left( \nwritesfn(\varnitems) \right) &< \biggo\left( \varnitems + \left( \frac{\growthfactor\varnitems - \initcapacity}{\growthfactor - 1} \right) \right)\\
	\biggo\left( \left( \frac{\growthfactor}{\growthfactor - 1} \right) \varnitems \right) \leq \biggo(\left( \nwritesfn(\varnitems) \right) &< \biggo\left( \left( \frac{2\growthfactor - 1}{\growthfactor - 1} \right) \varnitems \right)
	\end{align*}

@@

	\begin{align*}
	\spacefn(\varnitems) = \biggo(2^{\lceil \log_2 \varnitems \rceil + 1}) = \bigo(\varnitems)
	\end{align*}

@@

	I again start off by finding the write cost when $\varnitems$ are added. I give the write cost function a slightly different name, $\nwritesnewfn(\varnitems)$, so we can compare it to $\nwritesfn(\varnitems)$ later. I also use $\growthfactornew$ and $\initcapacitynew$ instead of $\growthfactor$ and $\initcapacity$ for the growth factor and initial capacity, respectively.

@@

	[lemma]
	The capacity of a funnel array grows by a constant factor $\growthfactornew$ after the first time it is called.

@@

	Let $k$ be the number of times the tail is added to. Since the tail is a dynamic array, the number of writes it makes is $\nwritesfn(k)$. Since we call [resize] $|S_R|$ times and it adds to the tail each time except for the first, $k = |S_R| - 1 = \lceil \log_{\growthfactornew} \varnitems - \log_{\growthfactornew} \initcapacitynew \rceil$. Then the formula for $\nwritesresizefn(n)$ is

@@

@@

	Indexing is another very common operation on a list. I will call methods that get or set an item at a specified index \textbf{get indexer} and \textit{set indexers}, respectively.

@@

	Given a strict ordering $<$, we say a list $\listname$ is \textbf{sorted} by $<$ iff $\left(a, b \in \listname \land a < b\right) \leftrightarrow I_M(a) < I_m(b)$. $I_M(a)$ is the last (maximum) index of $a$ in $L$, and $I_m(b)$ is the first (minimum) index of $b$ in $L$. Note that the use of $<$ on the right-hand side compares integers and not elements of $L$, so this is not a recursive definition.

>>>>

	Adding is the most common operation done on dynamic arrays\footnote{}. Funnel arrays improve the performance of adding in two ways: by allocating less memory, and reducing the amount of copying. I begin with the implementation of adding for dynamic arrays.

@@

	[proof]
	The capacity starts out at zero and becomes $\initcapacity$ once one element is added. From then on, it can only grow by a factor of $\growthfactor$ via [resize], so if $\growthfactor^i\initcapacity$ is the current capacity then $\growthfactor^{i + 1}\initcapacity$ must be the next capacity. By induction, the rest of the sequence is $$\left\{ \growthfactor^i\initcapacity \right\}_{i = 1}^k$$ for some $k$. Since the capacity is only as big as is needs to be, namely greater than or equal to $\varnitems$, $k$ is the smallest integer for which $\growthfactor^k\initcapacity \geq \varnitems$. Using this property, I now derive $k$.

@@

	[make lemma]
	Now, consider that to grow by $\growthfactor$ [resize] must be called. So $\growthfactor^k\initcapacity$ being present in $S_C$ means [resize] was called at size $\growthfactor^{k - 1}\initcapacity$ for $k \geq 1$. Then [resize] was called at the sizes $$S_R = 0,\ \initcapacity,\ \growthfactor\initcapacity,\ \growthfactor^2\initcapacity,\ \ldots\ \growthfactor^{\lceil \log_{\growthfactor} \varnitems - \log_{\growthfactor} \initcapacity \rceil - 1}\initcapacity$$

	Each time [resize] is called, $i$ items are copied where $i$ is the current size of the dynamic array. Then the total number of items copied for $\varnitems$ adds is

@@

	[ref2]
	To make this expression easier to work with, I will write it in big-P notation to approximate it for large $\varnitems$. First I note that $\log_{\growthfactor} k \leq \lceil \log_{\growthfactor} k \rceil < \left( \log_{\growthfactor} k \right) + 1$ for any $k > 0$, so $k \leq \growthfactor^{\lceil \log_{\growthfactor} k \rceil} < \growthfactor{k}$. Then

	\begin{align*}
	\nwritesfn(\varnitems) &= \varnitems + \left( \frac{\growthfactor^{\big\lceil \log_{\growthfactor} \left( \frac {\varnitems} {\initcapacity} \right) \big\rceil} - 1}{\growthfactor - 1} \right) \initcapacity\\
	\varnitems + \left( \frac{\frac{\varnitems}{\initcapacity} - 1}{\growthfactor - 1} \right) \initcapacity \leq \nwritesfn(\varnitems) &< \varnitems + \left( \frac{\growthfactor\frac{\varnitems}{\initcapacity} - 1}{\growthfactor - 1} \right) \initcapacity\\
	\varnitems + \left( \frac{\varnitems - \initcapacity}{\growthfactor - 1} \right) \leq \nwritesfn(\varnitems) &< \varnitems + \left( \frac{\growthfactor\varnitems - \initcapacity}{\growthfactor - 1} \right)\\
	\biggo\left( \varnitems + \left( \frac{\varnitems - \initcapacity}{\growthfactor - 1} \right) \right) \leq \biggo\left( \nwritesfn(\varnitems) \right) &< \biggo\left( \varnitems + \left( \frac{\growthfactor\varnitems - \initcapacity}{\growthfactor - 1} \right) \right)\\
	\biggo\left( \left( \frac{\growthfactor}{\growthfactor - 1} \right) \varnitems \right) \leq \biggo\left( \nwritesfn(\varnitems) \right) &< \biggo\left( \left( \frac{2\growthfactor - 1}{\growthfactor - 1} \right) \varnitems \right)
	\end{align*}

@@

	I wish to find the space allocated while adding $\varnitems$ items to a dynamic array. I denote this quantity $\spacefn(\varnitems)$, and define it as the total length of all arrays allocated during $\varnitems$ [add] calls. I now derive a formula for $\spacefn(\varnitems)$.

	From [lemma] I know the exact capacities $\listname$ takes on as $\varnitems$ items are added. Dynamic arrays also have the property that $i \in S_C$ iff an array of length $i$ was allocated. Then the total length of arrays allocated is

	\begin{align*}
	\spacefn(\varnitems) &= \sum_{i \in S_C} i\\
	&= 0 + \initcapacity + \growthfactor\initcapacity + \growthfactor^2\initcapacity + \ldots + \growthfactor^{\lceil \log_{\growthfactor} \varnitems - \log_{\growthfactor} \initcapacity \rceil}\initcapacity\\
	&= \left( \frac{\growthfactor^{\lceil \log_{\growthfactor} \varnitems - \log_{\growthfactor} \initcapacity \rceil + 1} - 1}{\growthfactor - 1} \right) \initcapacity
	\end{align*}

	Using the identity in [ref2] again, I bound its big-P complexity:

	\begin{align*}
	\left( \frac{\growthfactor \frac{\varnitems}{\initcapacity} - 1}{\growthfactor - 1} \right) \initcapacity \leq \spacefn(\varnitems) &< \left( \frac{\growthfactor^2 \frac{\varnitems}{\initcapacity} - 1}{\growthfactor - 1} \right) \initcapacity\\
	\biggo\left( \left( \frac{\growthfactor \frac{\varnitems}{\initcapacity} - 1}{\growthfactor - 1} \right) \initcapacity \right) \leq \biggo\left( \spacefn(\varnitems) \right) &< \biggo\left( \left( \frac{\growthfactor^2 \frac{\varnitems}{\initcapacity} - 1}{\growthfactor - 1} \right) \initcapacity \right)\\
	\biggo\left( \left( \frac{\growthfactor}{\growthfactor - 1} \right) \varnitems \right) \leq \biggo\left( \spacefn(\varnitems) \right) &< \biggo\left( \left( \frac{\growthfactor^2}{\growthfactor - 1} \right) \varnitems \right)
	\end{align*}

@@

	I again start off by finding the write cost when $\varnitems$ are added. I give the write cost function a slightly different name, $\nwritesnewfn(\varnitems)$, so I can compare it to $\nwritesfn(\varnitems)$ later. I also use $\growthfactornew$ and $\initcapacitynew$ instead of $\growthfactor$ and $\initcapacity$ for the growth factor and initial capacity, respectively.

@@

	[other lemma]
	The capacity of a funnel array grows by a constant factor $\growthfactornew$ after the first time it is called.

@@

	Let $k$ be the number of times the tail is added to. Since the tail is a dynamic array, the number of writes it makes is $\nwritesfn(k)$. Since [resize] is called $|S_R|$ times and it adds to the tail each time except for the first, $k = |S_R| - 1 = \lceil \log_{\growthfactornew} \varnitems - \log_{\growthfactornew} \initcapacitynew \rceil$. Then the formula for $\nwritesresizefn(n)$ is

@@

	% Should this be C(n) instead of C?
	The space function for $\varnitems$ [add] calls on a funnel array is denoted $\spacenewfn(\varnitems)$. Since funnel arrays never throw away item buffers, if the current capacity is $C$ then the total length of item buffers is also $C$. $\spacenewfn(\varnitems)$ is usually greater than $C$, however; this is because funnel arrays not only allocate item buffers, but stores references to them in the tail. Thus the space the tail allocates must also be determined.

	I determined in [...] that if $\varnitems$ are added then the tail is added to $k$ times, where $k = \lceil \log_{\growthfactornew} \varnitems - \log_{\growthfactornew} \initcapacitynew \rceil$. Since the tail is a dynamic array, it allocates $\spacefn(k)$ space, where $\spacefn$ is the space function of the tail. Through a similar method to [...] it can be shown $\bigo(\spacefn(k)) = \bigo(k) = \bigo(\log \varnitems)$. Then % TODO: P-analysis should come after the explicit formula.

@@

	Indexing is another very common operation on a list. I will call methods that get or set an item at a specified index \textbf{get} and \textbf{set indexers}, respectively.

@@

	Given a strict ordering $<$, I say a list $\listname$ is \textbf{sorted} by $<$ iff $\left(a, b \in \listname \land a < b\right) \leftrightarrow I_M(a) < I_m(b)$. $I_M(a)$ is the last (maximum) index of $a$ in $L$, and $I_m(b)$ is the first (minimum) index of $b$ in $L$. Note that the use of $<$ on the right-hand side compares integers and not elements of $L$, so this is not a recursive definition.
